Ecosystem carbon balance in the Hawaiian Islands under different scenarios of future climate and land use change

### 2021-05-14 Manuscript/HI_Model_ms.Rmd
updated manuscript files

## 1. Introduction
Terrestrial ecosystems are a major sink for atmospheric carbon dioxide (CO~2~), removing ~30% of human emissions on an annual basis and reducing the rate of increase in atmospheric CO~2~ [@keenan_terrestrial_2018; @friedlingstein_global_2019]. There is increasing recognition among policymakers that natural and agricultural ecosystems can contribute to climate mitigation, which has given rise to the popularity of “natural climate solutions” [@cameron_ecosystem_2017]. Defined as conservation and land management efforts aimed at enhancing ecosystem carbon storage [@griscom_natural_2017], natural climate solutions are appealing because they are seen as cost-efficient and readily available [@averchenkova_climate_2017; @cameron_ecosystem_2017; @fargione_natural_2018]. However, effective implementation is complicated by the uncertainty surrounding the future direction and magnitude of the land carbon sink, especially at the regional scale. Despite this uncertainty, evidence indicates that both interannual and long-term variability in carbon uptake by land ecosystems is driven primarily by fluctuations in climate, land use, and land cover change [@ahlstrom_dominant_2015; @prestele_current_2017; @friedlingstein_global_2019]. Incorporating the interactive effects of land use and climate into spatially explicit future projections of ecosystem carbon balance could therefore provide a reference point to evaluate the effectiveness of land-based mitigation. Although a complex challenge, the growing number of sub-national jurisdictions that plan to incorporate land-based mitigation strategies into their emissions reduction efforts would benefit from understanding how future land use and climate-biosphere feedbacks will affect ecosystem carbon balance in their respective regions [@sleeter_effects_2019].

@@

We used a stochastic, spatially explicit simulation model to estimate ecosystem carbon balance for Hawaiʻi's natural and agricultural lands on an annual basis for the period 2010–2100 under a range of assumptions about future climate, land use, land cover, disturbance, and global CO~2~ emissions [@daniel_state-and-transition_2016; @daniel_integrating_2018; @sleeter_effects_2019]. We explored four unique scenarios that represent all combinations of two land use change pathways (low and high) and two radiative forcing pathways (representative concentration pathway [RCP] 4.5 and RCP 8.5). In addition to these four scenarios, we conducted a separate series of simulations to examine how ecosystem carbon balance estimates vary in response to different levels of a CO~2~ fertilization effect (CFE) on net primary productivity [NPP; @sleeter_effects_2019]. our goals were to estimate changes in Hawaiian ecosystem carbon balance and their uncertainties under a range of plausible future scenarios, quantify the relative impact of major controlling processes such as land use change, disturbance, and climate change, and assess the sensitivity of model estimates to the introduction of a CFE on NPP.

## 2. Methods
We used the Land Use and Carbon Scenario Simulator (LUCAS), an integrated landscape change and carbon gain-loss model, to project changes in ecosystem carbon balance for the seven main Hawaiian Islands under all combinations of two land use scenarios (low and high) and two radiative forcing scenarios (RCP 4.5 and RCP 8.5). We also developed a separate set of scenarios to test model sensitivity to different levels of a CFE on NPP. The landscape change portion of LUCAS is a state-and-transition model that applies a Monte Carlo approach to track the state type and age of each simulation cell in response to a pre-determined set of transitions [@daniel_state-and-transition_2016]. The carbon gain-loss portion tracks carbon stocks within each simulation cell over time as continuous state variables, along with a pre-defined set of continuous flows specifying rates of change in stock levels over time [@daniel_integrating_2018; @sleeter_effects_2019]. We parameterized the Hawaiʻi LUCAS model to estimate annual changes in carbon stocks and fluxes in response to land use, land use change, wildland fire, and long-term climate variability for the time period 2010-2100.

@@

### *2.2 States and transitions*
We developed two land use scenarios (low and high) with transition pathways modified from @daniel_state-and-transition_2016. Transitions between state types were pre-defined to represent urbanization, agricultural contraction, agricultural expansion, harvesting of tree plantations, and wildfire. Agriculture, forest, grassland, shrubland, and tree plantation state types each had multiple transition pathways, while the barren state type could only transition to developed (i.e., urbanization). Although most state types had an urbanization transition pathway, there was no transition pathway out of an urbanized (developed) state. Water and wetland state types remained static throughout the simulation period.

 Transition targets were based on historical trends of land use change in the Hawaiian Islands from 1992-2011 [@noaa_coastal_2020] and on population projections for the State of Hawaiʻi [@kim_population_2018]. For the high land use scenario, transition rates for each timestep and Monte Carlo realization were sampled from uniform distributions bounded by the median and maximum historical rates of agricultural contraction, agricultural expansion, and urbanization for each island. For the low land use scenario, rates of agricultural contraction and expansion were sampled from uniform distributions bounded by zero and the minimum historical rates for each island. Urbanization rates in the low land use scenario were based on island-level population estimates and projections at five year intervals from 2010-2045 [@kim_population_2018]. We converted population projections into urbanization transition targets following @sleeter_future_2017 by calculating population density for each island and then projecting future developed area based on the five-year incremental change in island population. The spatial extent of agricultural contraction, agricultural expansion, and urbanization was constrained in both land use scenarios based on existing zoning maps [@daniel_state-and-transition_2016]. Transition targets for tree plantation harvest were set at ~75% of recent historical rates in the high land use scenario and ~40% of recent historical rates in the low land use scenario [@daniel_state-and-transition_2016]. In both land use scenarios, approximately 60% of tree plantation harvests were replacement harvests resulting in conversion to agriculture. The remaining 40% were rotation harvests replanted to *Eucalyptus* spp.

@@

 Net primary production for each simulation cell was calculated as the mean IBIS-derived value for each combination of moisture zone and state type adjusted with a spatially explicit stationary growth multiplier to reflect local variation driven by microclimate [@sleeter_effects_2019]. We calculated this spatial growth multiplier as the NPP anomaly for each simulation cell relative to mean NPP values for each combination of moisture zone and state type based on empirical relationships between total annual NPP and mean annual rainfall or temperature [@schuur_productivity_2003; @del_grosso_global_2008] using Hawaiʻi-specific climate data [@giambelluca_online_2013; @giambelluca_evapotranspiration_2014]. Soil carbon flux to the atmosphere (R~h~) and aquatic soil carbon losses (leaching and overland flow) were estimated as the ratio of the IBIS-derived flux for each combination of moisture zone and state type to the microclimate-adjusted NPP value for each simulation cell. All other carbon flow rates were estimated as the ratio of the mean IBIS-derived flux for each combination of moisture zone and state type to the size of the originating carbon stock at each age [@sleeter_effects_2018; @daniel_integrating_2018]. Climate change impacts on carbon flows were represented by temporal growth and decay multipliers applied to each simulation cell based on mid-century (2049-2069) and end-of-century (2070-2099) statistically downscaled CMIP5 temperature and rainfall projections for the Hawaiian Islands under each of two radiative forcing scenarios [RCP 4.5 and RCP 8.5; @timm_statistical_2015; @timm_future_2017]. The impact of future changes in rainfall and temperature on NPP were represented by annual growth multipliers calculated using empirical NPP models [@schuur_productivity_2003; @del_grosso_global_2008] and climate model projections of temperature and rainfall for each radiative forcing scenario. The effect of future warming on turnover rates of dead organic matter were represented by temporal decay multipliers calculated using Q10 functions and climate model temperature projections for each radiative forcing scenario. We applied a Q10 of 2.0 for wood and soil organic matter decay flows [@kurz_cbm-cfs3:_2009; @sleeter_effects_2019] and a Q10 of 2.17 for litter decay flows [@bothwell_leaf_2014]. Transition-triggered carbon flows resulting from disturbances associated with land use change, timber harvesting, and wildfire were based on values from @don_impact_2011, @selmants_baseline_2017, and @daniel_integrating_2018.

### *2.4 CO~2~ fertilization effect*
Increasing atmospheric CO~2~ concentrations stimulate leaf-level photosynthesis, potentially increasing NPP as well [@franks_sensitivity_2013]. However, the magnitude and persistence of this effect is highly uncertain, particularly across a range of climatic conditions and over long time spans [@walker_integrating_2020]. Following @sleeter_effects_2019, we developed a separate set of scenarios designed to test the sensitivity of LUCAS model projections of ecosystem carbon balance to different rates of a CO~2~ fertilization effect (CFE). We incorporated a CFE multiplier for NPP that represented the percent increase in NPP for every 100 ppm increase in atmospheric CO~2~ concentration under the high land use and high radiative forcing (RCP 8.5) scenario. We tested five CFE levels ranging from 5% to 15%, which is within the range of CFEs observed in free air CO~2~ enrichment (FACE) experiments. For all levels, we assumed CFEs reached saturation at an atmospheric CO~2~ concentration of 600 ppm, with no further stimulation of NPP despite a continued increase in CO~2~ concentration to 930 ppm by 2100. This 600ppm threshold generally coincides with the upper limit from FACE experiments and is reached by the year 2060 under RCP 8.5.

@@

Interannual climate variability is a primary factor influencing spatial and temporal patterns of global wildland fire activity [@abatzoglou_global_2018], with climate warming expected to increase wildland fire frequency and wildfire season length across a wide range of biomes [@sun_global_2019]. Although our model projections capture the spatial and temporal variation in ignition probability and fire extent by sampling from previous fire years (1999-2019), we did not incorporate how projected future climate change will affect the frequency and extent of wildland fire. Recent fire probability modeling for the northwest portion of Hawaiʻi Island indicated that projected drying and warming trends under RCP 8.5 could increase maximum fire probability values more than three-fold and shift areas of peak flammability to higher elevation by mid-century [@trauernicht_vegetationrainfall_2019]. Extending this probability fire modeling approach statewide would provide a quantitative, spatially explicit assessment of wildland fire probability for the main Hawaiian Islands as predicted by climate, land cover, and ignition density, which is highly correlated with population density [@trauernicht_contemporary_2015; @trauernicht_vegetationrainfall_2019]. This approach would provide future simulation model projections of Hawaiian ecosystem carbon balance with more realistic scenarios of expected annual area burned based on the integrated effects of future climate and land use change.

>>>>

## 1. Introduction
Terrestrial ecosystems are a major sink for atmospheric carbon dioxide (CO~2~), removing ~30% of human emissions on an annual basis and reducing the rate of increase in atmospheric CO~2~ [@keenan_terrestrial_2018; @friedlingstein_global_2019]. There is increasing recognition among policymakers that natural and agricultural ecosystems can contribute to climate mitigation, which has given rise to the popularity of “natural climate solutions” [@cameron_ecosystem_2017]. Defined as conservation and land management efforts aimed at enhancing ecosystem carbon storage [@griscom_natural_2017], natural climate solutions are appealing because they are seen as cost-efficient and readily available [@averchenkova_climate_2017; @cameron_ecosystem_2017; @fargione_natural_2018]. However, effective implementation is complicated by the uncertainty surrounding the future direction and magnitude of the land carbon sink, especially at the regional scale. Despite this uncertainty, evidence indicates that both interannual and long-term variability in carbon uptake by land ecosystems is driven primarily by fluctuations in climate, land use, and land cover change [@ahlstrom_dominant_2015; @prestele_current_2017; @friedlingstein_global_2019]. Incorporating the interactive effects of land use and climate into spatially explicit future projections of ecosystem carbon balance could therefore provide a reference point to evaluate the effectiveness of land-based mitigation. Although a complex challenge, a growing number of sub-national jurisdictions plan to incorporate land-based mitigation strategies into their emissions reduction efforts. These jurisdictions would benefit from understanding how future land use and climate-biosphere feedbacks will affect ecosystem carbon balance in their respective regions [@sleeter_effects_2019].

@@

We used a stochastic, spatially explicit simulation model to estimate ecosystem carbon balance for Hawaiʻi's natural and agricultural lands on an annual basis for the period 2010–2100 under a range of assumptions about future climate, land use, land cover, disturbance, and global CO~2~ emissions [@daniel_state-and-transition_2016; @daniel_integrating_2018; @sleeter_effects_2019]. We explored four unique scenarios that represent all combinations of two land use change pathways (low and high) and two radiative forcing pathways (representative concentration pathway [RCP] 4.5 and RCP 8.5). In addition to these four scenarios, we conducted a separate series of simulations to examine how ecosystem carbon balance estimates vary in response to different levels of a CO~2~ fertilization effect (CFE) on net primary productivity [NPP; @sleeter_effects_2019]. Our goals were to estimate changes in Hawaiian ecosystem carbon balance and their uncertainties under a range of plausible future scenarios, quantify the relative impact of major controlling processes such as land use change, disturbance, and climate change, and assess the sensitivity of model estimates to the introduction of a CFE on NPP.

## 2. Methods
We used the Land Use and Carbon Scenario Simulator (LUCAS), an integrated landscape change and carbon gain-loss model, to project changes in ecosystem carbon balance for the seven main Hawaiian Islands. The landscape change portion of LUCAS is a state-and-transition model, where 'state' is the pre-defined land cover class  assigned to each simulation cell and 'transition' is the change in state of each simulation cell over time [@daniel_state-and-transition_2016]. Landscape change occurs by applying a Monte Carlo approach to track the state type and age of each simulation cell in response to a pre-determined set of transition pathways [@daniel_state-and-transition_2016]. The carbon gain-loss portion tracks carbon stocks within each simulation cell over time as continuous state variables, along with a pre-defined set of continuous flows specifying rates of change in stock levels over time [@daniel_integrating_2018; @sleeter_effects_2019]. We parameterized the Hawaiʻi LUCAS model to estimate annual changes in carbon stocks and fluxes in response to land use, land use change, wildland fire, and long-term climate variability for the time period 2010-2100.

@@

### *2.2 States and transitions*
We developed two land use scenarios (low and high) with transition pathways modified from @daniel_state-and-transition_2016. Transition pathways between state types were pre-defined to represent urbanization, agricultural contraction, agricultural expansion, harvesting of tree plantations, and wildfire. Agriculture, forest, grassland, shrubland, and tree plantation state types each had multiple transition pathways, while the barren state type could only transition to developed (i.e., urbanization). Although most state types had an urbanization transition pathway, there was no transition pathway out of an urbanized (developed) state. Water and wetland state types remained static throughout the simulation period.

 Transition targets set the area to be transitioned over time. These targets were based on historical trends of land use change in the Hawaiian Islands from 1992-2011 [@noaa_coastal_2020] and on population projections for the State of Hawaiʻi [@kim_population_2018]. For the high land use scenario, transition rates for each timestep and Monte Carlo realization were sampled from uniform distributions bounded by the median and maximum historical rates of agricultural contraction, agricultural expansion, and urbanization for each island. For the low land use scenario, rates of agricultural contraction and expansion were sampled from uniform distributions bounded by zero and the minimum historical rates for each island. Urbanization rates in the low land use scenario were based on island-level population estimates and projections at five year intervals from 2010-2045 [@kim_population_2018]. We converted population projections into urbanization transition targets following @sleeter_future_2017 by calculating population density for each island and then projecting future developed area based on the five-year incremental change in island population. The spatial extent of agricultural contraction, agricultural expansion, and urbanization was constrained in both land use scenarios based on existing zoning maps [@daniel_state-and-transition_2016]. Transition targets for tree plantation harvest were set at ~75% of recent historical rates in the high land use scenario and ~40% of recent historical rates in the low land use scenario [@daniel_state-and-transition_2016]. In both land use scenarios, approximately 60% of tree plantation harvests were replacement harvests resulting in conversion to agriculture. The remaining 40% were rotation harvests replanted to *Eucalyptus* spp.

@@

 Net primary production for each simulation cell was calculated as the mean IBIS-derived value for each combination of moisture zone and state type adjusted with a spatially explicit stationary growth multiplier to reflect local variation driven by microclimate [@sleeter_effects_2019]. We calculated this spatial growth multiplier as the NPP anomaly for each simulation cell relative to mean NPP values for each combination of moisture zone and state type based on empirical relationships between total annual NPP and mean annual rainfall or temperature [@schuur_productivity_2003; @del_grosso_global_2008] using Hawaiʻi-specific climate data [@giambelluca_online_2013; @giambelluca_evapotranspiration_2014]. Soil carbon flux to the atmosphere (R~h~) and aquatic soil carbon losses (leaching and overland flow) were estimated as the ratio of the IBIS-derived flux for each combination of moisture zone and state type to the microclimate-adjusted NPP value for each simulation cell. All other carbon flow rates were estimated as the ratio of the mean IBIS-derived flux for each combination of moisture zone and state type to the size of the originating carbon stock at each age [@sleeter_effects_2018; @daniel_integrating_2018].

 Climate change impacts on carbon flows were represented by temporal growth and decay multipliers applied to each simulation cell based on mid-century (2049-2069) and end-of-century (2070-2099) statistically downscaled CMIP5 temperature and rainfall projections for the Hawaiian Islands under each of two radiative forcing scenarios [RCP 4.5 and RCP 8.5; @timm_statistical_2015; @timm_future_2017]. The impact of future changes in rainfall and temperature on NPP were represented by annual growth multipliers calculated using empirical NPP models [@schuur_productivity_2003; @del_grosso_global_2008] and climate model projections of temperature and rainfall for each radiative forcing scenario. The effect of future warming on turnover rates of dead organic matter were represented by temporal decay multipliers calculated using Q~10~ temperature coefficients based on climate model temperature projections for each radiative forcing scenario. The Q~10~ temperature coefficients represented the proportional change in detrital carbon turnover due to a 10° C change in mean annual temperature. We applied a Q~10~ of 2.0 for wood and soil organic matter decay flows [@kurz_cbm-cfs3:_2009; @sleeter_effects_2019] and a Q~10~ of 2.17 for litter decay flows [@bothwell_leaf_2014]. Transition-triggered carbon flows resulting from disturbances associated with land use change, timber harvesting, and wildfire were based on values from @don_impact_2011, @selmants_baseline_2017, and @daniel_integrating_2018.

### *2.4 CO~2~ fertilization effect*
Increasing atmospheric CO~2~ concentrations stimulate leaf-level photosynthesis, potentially increasing NPP as well [@franks_sensitivity_2013]. However, the magnitude and persistence of this effect is highly uncertain, particularly across a range of climatic conditions and over long time spans [@walker_integrating_2020]. Following @sleeter_effects_2019, we developed a separate set of scenarios designed to test the sensitivity of LUCAS model projections of ecosystem carbon balance to different rates of a CO~2~ fertilization effect (CFE). We incorporated a CFE multiplier for NPP that represented the percent increase in NPP for every 100 ppm increase in atmospheric CO~2~ concentration under the high land use and high radiative forcing (RCP 8.5) scenario. We tested five CFE levels ranging from 5% to 15%, which is within the range of CFEs observed in free air CO~2~ enrichment (FACE) experiments. For all levels, we assumed CFEs reached saturation at an atmospheric CO~2~ concentration of 600 ppm, with no further stimulation of NPP despite a continued increase in CO~2~ concentration to 930 ppm by 2100. This 600 ppm threshold generally coincides with the upper limit from FACE experiments and is reached by the year 2060 under RCP 8.5.

@@

Interannual climate variability is a primary factor influencing spatial and temporal patterns of global wildland fire activity [@abatzoglou_global_2018], with climate warming expected to increase wildland fire frequency and wildfire season length across a wide range of biomes [@sun_global_2019]. Although our model projections capture the spatial and temporal variation in ignition probability and fire extent by sampling from previous fire years (1999-2019), we did not incorporate the effects of projected future climate change on the frequency and extent of wildland fire. Recent fire probability modeling for the northwest portion of Hawaiʻi Island indicated that projected drying and warming trends under RCP 8.5 could increase maximum fire probability values more than three-fold and shift areas of peak flammability to higher elevation by mid-century [@trauernicht_vegetationrainfall_2019]. Extending this probability fire modeling approach statewide would provide a quantitative, spatially explicit assessment of wildland fire probability for the main Hawaiian Islands as predicted by climate, land cover, and ignition density, which is highly correlated with population density [@trauernicht_contemporary_2015; @trauernicht_vegetationrainfall_2019]. This approach would provide future simulation model projections of Hawaiian ecosystem carbon balance with more realistic scenarios of expected annual area burned based on the integrated effects of future climate and land use change.
